Analysis v1.2 mash

### 2023-09-12 R/Analysis_v1.2_mash_using_greedy_covar.Rmd
analysis: Substitute h2 for CV on diagonals of hypothesis covar matrices

##

# ----
## load phe & wea data
```{r}
phe_ten_sites <- readRDS(here("data", "Phenotypes_2018_to_2021_10_sites.rds"))
jday_phe <- readRDS("~/Github/pvdiv-phenology-gxe/data/Julian_date_GDD_and_rainfall_phenotypes_4wcr_and_gwas.rds")
wea_ten_sites <- readRDS(file = here("data", "Daily_weather_2018_to_2021_10_sites.rds"))

@@

# Covar csv for greedy mash algorithm

@@

@@

fl_ten_sites <- fl_ten_sites %>%
  left_join(select(metadata, PLANT_ID, SUBPOP)) %>%
  rename(Year = YEAR)

@@

phe_hyp <- colnames(fl_ten_sites)[c(5,9:51)]
i=1
k=1

@@

    cov_df <- fl_ten_sites %>%
      ungroup() %>%
      filter(SUBPOP %in% subpop_v2[[k]]) %>%
      filter(Year %in% 2019) %>%
      select(PLANT_ID, SUBPOP, manu_site, phe_hyp[i]) %>%
      group_by(PLANT_ID, SUBPOP, manu_site) %>%
      mutate(IND = row_number()) %>%
      pivot_wider(names_from = manu_site, values_from = phe_hyp[i]) %>%
      mutate(PLANT_ID = as.factor(PLANT_ID),
             IND = as.factor(IND)) %>%
      select(PLANT_ID, SUBPOP, IND, MI, MO, NE, OK, SD, TX1, TX2, TX3) #####  IL,

    cor_phe <- cor(cov_df[,-(1:3)], use = "pairwise")
    # Set diagonal of this matrix = coefficient of variation within each garden.
    diag(cor_phe) <- cov_df %>%
      ungroup() %>%
      summarise(#IL = sqrt(var(IL, na.rm = TRUE))/mean(IL, na.rm = TRUE),  #####
                MI = sqrt(var(MI, na.rm = TRUE))/mean(MI, na.rm = TRUE),  #####
                MO = sqrt(var(MO, na.rm = TRUE))/mean(MO, na.rm = TRUE),
                NE = sqrt(var(NE, na.rm = TRUE))/mean(NE, na.rm = TRUE),
                OK = sqrt(var(OK, na.rm = TRUE))/mean(OK, na.rm = TRUE),
                SD = sqrt(var(SD, na.rm = TRUE))/mean(SD, na.rm = TRUE),
                TX1 = sqrt(var(TX1, na.rm = TRUE))/mean(TX1, na.rm = TRUE),
                TX2 = sqrt(var(TX2, na.rm = TRUE))/mean(TX2, na.rm = TRUE),
                TX3 = sqrt(var(TX3, na.rm = TRUE))/mean(TX3, na.rm = TRUE)) %>%
      as_vector()

    U_hyp_2019 <- list.prepend(U_hyp_2019, cor_phe)
    names(U_hyp_2019)[1] <- paste0("flowering.", names(subpop_v2)[k], ".", phe_hyp[i], ".U.mat")
  }
}

U_hyp_2019
length(U_hyp_2019)
for (i in 1:length(U_hyp_2019)) {
  d_out <- as.data.frame(U_hyp_2019[[i]], optional = FALSE) %>%
    rownames_to_column(var = " ")
  write_csv(d_out, file = here("data", "greedy_more_cov", "hypothesis_matrices", paste0(names(U_hyp_2019)[i], ".csv")))
}
```

@@

    cor_phe <- cor(cov_df[,-(1:3)], use = "pairwise")
    # Set diagonal of this matrix = coefficient of variation within each garden.
    diag(cor_phe) <- cov_df %>%
      ungroup() %>%
      summarise(#IL = sqrt(var(IL, na.rm = TRUE))/mean(IL, na.rm = TRUE),  #####
                MI = sqrt(var(MI, na.rm = TRUE))/mean(MI, na.rm = TRUE),  #####
                MO = sqrt(var(MO, na.rm = TRUE))/mean(MO, na.rm = TRUE),
                NE = sqrt(var(NE, na.rm = TRUE))/mean(NE, na.rm = TRUE),
                OK = sqrt(var(OK, na.rm = TRUE))/mean(OK, na.rm = TRUE),
                SD = sqrt(var(SD, na.rm = TRUE))/mean(SD, na.rm = TRUE),
                TX1 = sqrt(var(TX1, na.rm = TRUE))/mean(TX1, na.rm = TRUE),
                TX2 = sqrt(var(TX2, na.rm = TRUE))/mean(TX2, na.rm = TRUE),
                TX3 = sqrt(var(TX3, na.rm = TRUE))/mean(TX3, na.rm = TRUE)) %>%
      as_vector()

    U_hyp_2019 <- list.prepend(U_hyp_2019, cor_phe)
    names(U_hyp_2019)[1] <- paste0("greenup.", names(subpop_v2)[k], ".", phe_hyp[i], ".U.mat")
  }
}
for (i in 1:length(U_hyp_2019)) {
  d_out <- as.data.frame(U_hyp_2019[[i]], optional = FALSE) %>%
    rownames_to_column(var = " ")
  write_csv(d_out, file = here("data", "greedy_more_cov", "hypothesis_matrices", paste0(names(U_hyp_2019)[i], ".csv")), )
}
```

@@

>>>>

# ----

# Covar 1-20d weather
As of 2023-09-11, want to remake these per reviewer revision to have h2 on the diagonal instead of the CV. Need to estimate h2 for these new covar matrices, probably using rrBLUP.

## load phe & wea data
```{r}
phe_ten_sites <- readRDS(here("data", "Phenotypes_2018_to_2021_10_sites.rds"))
sites <- readRDS("~/Github/pvdiv-phenology-gxe/data/sites.rds")
jday_phe <- readRDS("~/Github/pvdiv-phenology-gxe/data/Julian_date_GDD_and_rainfall_phenotypes_4wcr_and_gwas.rds")
wea_ten_sites <- readRDS(file = here("data", "Daily_weather_2018_to_2021_10_sites.rds"))

@@

# ---
# Covar csv for greedy mash algorithm

@@

## Calculate heritability for diagonal
I don't have access to ASREML any more, so need a different package to do this.
```{r}
library(rrBLUP)
k_full <- read_rds("~/Github/pvdiv-genome/tensite_twoyear/Kinship_van_Raden_630_individuals_SNPs_r2_20percent.rds")
fl_ten_sites
colnames(k_full)

phe_one <- fl_ten_sites %>% filter(manu_site == "TX1") %>%
  filter(WHERE != "FWCR") %>%
  group_by(PLANT_ID) %>%
  summarise(crain_1d = mean(crain_1d, na.rm = TRUE)) %>%
  filter(!is.na(PLANT_ID))
k_in_phe <- which(colnames(k_full) %in% phe_one$PLANT_ID)
k_one <- k_full[k_in_phe, k_in_phe]

nrow(k_one)
nrow(phe_one)
mod2 <- mixed.solve(y = phe_one$crain_1d, K = k_one)
h2 <- mod2$Vu/(mod2$Vu + mod2$Ve)
```

@@

fl_ten_sites <- fl_ten_sites %>%
  left_join(select(metadata, PLANT_ID, SUBPOP)) %>%
  rename(Year = YEAR)
fl_ten_sites$PLANT_ID <- factor(fl_ten_sites$PLANT_ID, levels = colnames(k_full))

@@

phe_hyp <- colnames(fl_ten_sites)[c(5,9:51)]
i=2
k=1

@@

    cov_df <- fl_ten_sites %>%
      ungroup() %>%
      filter(SUBPOP %in% subpop_v2[[k]]) %>%
      filter(Year %in% 2019) %>%
      select(PLANT_ID, SUBPOP, manu_site, phe_hyp[i]) %>%
      group_by(PLANT_ID, SUBPOP, manu_site) %>%
      mutate(IND = row_number()) %>%
      pivot_wider(names_from = manu_site, values_from = phe_hyp[i]) %>%
      mutate(PLANT_ID = as.factor(PLANT_ID),
             IND = as.factor(IND)) %>%
      select(PLANT_ID, SUBPOP, IND, MI, MO, NE, OK, SD, TX1, TX2, TX3) %>% #####  IL,
      replace_na(list(MI = 0, MO = 0, NE = 0, OK = 0, SD = 0, TX1 = 0, TX2 = 0, 
                      TX3 = 0))

    cor_phe <- cor(cov_df[,-(1:3)], use = "pairwise")
    # Set diagonal of this matrix = heritability within each garden
    h2_diag <- c()
    site_names <- c("MI", "MO", "NE", "OK", "SD", "TX1", "TX2", "TX3")
    for(j in seq_along(site_names)) {
      phe_one <- fl_ten_sites %>% filter(manu_site == site_names[j]) %>%
        filter(WHERE != "FWCR") %>%
        group_by(PLANT_ID) %>%
        summarise(phenotype = mean(!! sym(phe_hyp[i]), na.rm = TRUE)) %>%
        filter(!is.na(PLANT_ID))
      k_in_phe <- which(colnames(k_full) %in% phe_one$PLANT_ID)
      k_one <- k_full[k_in_phe, k_in_phe]
      mod2 <- mixed.solve(y = phe_one$phenotype, K = k_one)
      h2_diag[j] <- mod2$Vu/(mod2$Vu + mod2$Ve)
    }

    diag(cor_phe) <- h2_diag
    if (!matrixcalc::is.positive.semi.definite(cor_phe)) {
      cor_phe <- Matrix::nearPD(cor_phe)
    }
    U_hyp_2019 <- list.prepend(U_hyp_2019, cor_phe)
    names(U_hyp_2019)[1] <- paste0("flowering.", names(subpop_v2)[k], ".", 
                                   phe_hyp[i], ".U.mat")
  }
}

U_hyp_2019

# Test that every matrix is positive semi definite. Should all be true now changed code above.
U_psd <- c() 
for (i in seq_along(U_hyp_2019)) {
  if (is.matrix(U_hyp_2019[[i]])) {
    U_psd[i] <- matrixcalc::is.positive.semi.definite(U_hyp_2019[[i]])
  } else {
    U_psd[i] <- matrixcalc::is.positive.semi.definite(as.matrix(U_hyp_2019[[i]]$mat))
  }

for (i in seq_along(U_hyp_2019)) {
  if (is.matrix(U_hyp_2019[[i]])) {
    matrixout = matrix(unlist(U_hyp_2019[[i]]), ncol = 8, byrow = TRUE)
  } else {
    matrixout = matrix(unlist(U_hyp_2019[[i]]$mat), ncol = 8, byrow = TRUE)
  }
  row.names(matrixout) = site_names
  colnames(matrixout) = site_names
  #print(matrixout)
  write.csv(matrixout, file = here("analysis", "mash_greedy_algorithm", 
                                   "mash_h2_diag_covar", "matrices", 
                                   paste0(names(U_hyp_2019)[i], ".csv")))
}

```
Another error: "Error in mash(data, covmats) : 
  All U_scaled matrices should be positive semi-definite

@@

    cor_phe <- cor(cov_df[,-(1:3)], use = "pairwise")
    # Set diagonal of this matrix = heritability within each garden
    h2_diag <- c()
    site_names <- c("MI", "MO", "NE", "OK", "SD", "TX1", "TX2", "TX3")
    for(j in seq_along(site_names)) {
      phe_one <- fl_ten_sites %>% filter(manu_site == site_names[j]) %>%
        filter(WHERE != "FWCR") %>%
        group_by(PLANT_ID) %>%
        summarise(phenotype = mean(!! sym(phe_hyp[i]), na.rm = TRUE)) %>%
        filter(!is.na(PLANT_ID))
      k_in_phe <- which(colnames(k_full) %in% phe_one$PLANT_ID)
      k_one <- k_full[k_in_phe, k_in_phe]
      mod2 <- mixed.solve(y = phe_one$phenotype, K = k_one)
      h2_diag[j] <- mod2$Vu/(mod2$Vu + mod2$Ve)
    }

    diag(cor_phe) <- h2_diag
    if (!matrixcalc::is.positive.semi.definite(cor_phe)) {
      cor_phe <- Matrix::nearPD(cor_phe)
    }

    U_hyp_2019 <- list.prepend(U_hyp_2019, cor_phe)
    names(U_hyp_2019)[1] <- paste0("greenup.", names(subpop_v2)[k], ".", phe_hyp[i], ".U.mat")
  }
}

for (i in seq_along(U_hyp_2019)) {
  if (is.matrix(U_hyp_2019[[i]])) {
    matrixout = matrix(unlist(U_hyp_2019[[i]]), ncol = 8, byrow = TRUE)
  } else {
    matrixout = matrix(unlist(U_hyp_2019[[i]]$mat), ncol = 8, byrow = TRUE)
  }
  row.names(matrixout) = site_names
  colnames(matrixout) = site_names
  #print(matrixout)
  write.csv(matrixout, file = here("analysis", "mash_greedy_algorithm", 
                                   "mash_h2_diag_covar", "matrices", 
                                   paste0(names(U_hyp_2019)[i], ".csv")))
}

@@

# ---------------
# Info to run greedy mash algo

data/mash_switchgrass_models/mash/all_matrices/matrices hosts all covar mat (?)
analysis/mash_greedy_algorithm/mash_switchgrass_models/mash/all_matrices/matrices has more? covar mat?

Set up new analysis/mash_greedy_algorithm/mash_1_to_28d_covar/ or subdirectory to have all the subdir necessary to run `run.mash.greedy.search.py` python script.

Make sure all matrices are present in matrices/ subdir and named appropriately for recognition
Make sure random and strong Bhat and Shat in combined_mash_effects and combined_mash_stderrs are appropriate - perhaps replace these with my inputs in mash/Subpop/inputs; perhaps use what's here as tests first

May need to adjust midwest covar & effects appropriately to reflect only 6 (floewring) and 7 (greenup) rows & columns used; other GWAS were dropped & I think including them would be misleading, likely.
